\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-16 phd-cs-us.tex
updates

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as \emph{Vietnamese}, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to give some details about the PhD Admission process and share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is written for Vietnamese students interested in Computer Science, it should apply to students in many other countries and other majors.
Moreover, while many examples given are for George Mason University, the writing should generalize to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

In addition, this document can help \emph{US faculty and admission committee members} understand more about international students, e.g., questions, concerns, and various cultural differences.  By understanding, appreciating, and even leveraging these information, CS programs in the US can attract larger and stronger application pools from Vietnam and other countries.

I wish you the best of luck. And if you follow these advice,
you will at least have a good chance at GMU (see
\href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU}{why
you want to study at GMU}). Happy school hunting!

@@

After you submit your PhD application (usually in December), it will be first screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters?
Then your application will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS. Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and try to reach a consensus about your case.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can give inputs on your profile.

@@

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} since a PhD in Computer Science
is a research degree. To assess your research capability, we consider
the following key indicators, listed in order of importance.

@@

\begin{wrapfigure}{r}{0.50\textwidth}
\vspace{-0.4in}
  \begin{center}
    \includegraphics[width=0.50\textwidth]{c2.png}
  \end{center}
\vspace{-0.4in}  
\end{wrapfigure}
Finally, this is something easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? talk about a couple of professors who you're interested in working with (in many cases your application will be forwarded to them for evaluation).
This shows that you're serious and have done homework on places you're applying to.
Be careful not to send statements to wrong schools or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.

\begin{commentbox}[Vu]
I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\end{commentbox}

@@

An interview lasts about 15--30 mins, and one implicit thing you will be evaluated on is whether you can communicate effectively (i.e., speak/understand English).  You will also get chance to ask questions about the university so think of something to ask (just the same as you interview at a company).

\begin{commentbox}[Vu]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and wants argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{commentbox}

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you didn't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.
\begin{commentbox}[Dat] Most Vietnamese students, including those from top schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?}

\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.
Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured) and expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research, e.g., they are more involved in administrative duties.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
    \vspace{-0.4in}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
    Typically, 5-7 years for PhD in CS at US universities.  The first 1.5--2 years you spend on coursework and learning research.  The next 2--3 years you focus on your research, formi dissertation topic, and get results published.  The last 1-2 years you write and defend your dissertation and look for job. The PhDComics figure on the right shows the "ambition" level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is "Please just get me graduate").

@@

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for funding).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student makes two conference trips per year, one domestic and one international (so conf. registration, airline tickets, taxi, meals, etc are all included). I do not include budgets such as computer equipments. At the end, the total budget comes out to be \$69K.  \emph{The summary is that while you're paid X, your advisor probably pays 2X for you}.

>>>>

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as \emph{Vietnam}, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is intended for Vietnamese students interested in Computer Science, it may be applicable to students in from various countries and disciplines.
Moreover, although many examples are for George Mason University, the writing should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

In addition, this document can help \emph{US faculty and admission committee} understand more about international students and their cultural differences.  By understanding, appreciating, and even leveraging these differences, CS programs in the US can attract larger and more competitive application pools from Vietnam and other countries.

I wish you the best of luck. And if you follow this guideline,
you will at least have a good chance at GMU (see
\href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU}{why
you want to study at GMU}). Happy school hunting!

@@

After you submit your PhD application (usually in December), it will be first screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters?

Application will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size of the committee depends on the department size, e.g., around 15 at GMU. Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can provide inputs and opinions on your profile.

@@

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} since a PhD in Computer Science
is a research degree and determine if you would \textbf{fit} the program. To evaluate your profile, we consider
the following key indicators, listed in order of importance.

@@

\begin{wrapfigure}{r}{0.52\textwidth}
\vspace{-0.35in}
  \begin{center}
    \includegraphics[width=0.52\textwidth]{c2.png}
  \end{center}
\vspace{-0.4in}  
\end{wrapfigure}
Finally, this is something easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? mention the names of professors who you're interested in (in many cases your application will be forwarded to them for evaluation).
This shows that you're serious and have done homework on places you're applying to.
Admission committee will look for this part at the end of the statement, so do not skip it.

\begin{commentbox}[Vu]
I always read the research statement first and then LORs. If I am
persuaded by then, I would skim over other factors and advocate for
admission (unless I see red flags in other parts). If I am not
convinced, then I will likely recommend rejection (unless I see
something standout in other parts).
\tcblower
Be careful not to send statements to wrong schools or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.
\end{commentbox}

@@

An interview lasts about 15--30 mins, and one implicit thing you will be evaluated on is whether you can communicate effectively (i.e., speak/understand English).  You will also get chance to ask questions about the university so think of something to ask (just the same as interviewing at a company).

\begin{commentbox}[Vu]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and wants argue for admission, i.e., without the interview, that application is likely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{commentbox}

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you didn't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

\begin{commentbox}[Dat] Most Vietnamese students, including those from top schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\subsection{Tenured or tenure-track faculty? Who do you choose?}
\begin{commentbox}[Thanh]
In my opinion, having a well-suited advisor is crucial for a successful PhD and research career. One effective approach to finding a suitable professor is by working with a professor during your undergraduate studies. An exemplary instance is VinAI's residency program, where residents collaborate with professors from the US for two years before applying to PhD programs. Many VinAI residents have achieved remarkable results and gained admission to prestigious US universities. Unfortunately, VinAI's resident program is limited to AI research. In other fields, e.g. SE,
Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
\end{commentbox}
\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

Tenured faculty (e.g., associate and full profs.) are more likely to be older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured) and expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research, e.g., they are more involved in administrative duties.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
    \vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
    Typically, 5-7 years for PhD in CS at US universities.  The first 2 years you spend on coursework, finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published.  The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. The PhDComics figure on the right shows the "ambition" level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is "Just get me graduate").

@@

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for funding).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student makes two conference trips per year, one domestic and one international (so conf. registration, airline tickets, taxi, meals, etc are all included). At the end, the total budget comes out to be \$69K. \emph{The summary is that while you're paid X, your advisor probably pays 2X for you}.
